---
title: "ROSS Cache La Poudre (CLP) Water Quality Data"
format: html
  self-contained: true
code-fold: true
warning: false
editor: visual
---

# ROSS Cache La Poudre (CLP) Water Quality Data

## Before running:

- Ensure new raw file is added to `data/raw/` folder and the name is updated in the `newest_raw_file` variable below.
- Ensure no samples need to be removed due to lab QA or sampling issues (e.g. bottle contamination, etc.). If so, remove them from the `water_chemistry_import` data frame below.
- Ensure new meas file is added to `data/raw/` folder and the name is updated in the `newest_meas_file` variable below.
- If new site is sampled, add to `location_metadata.csv` and ensure location names and coordinates are correct/match existing sites

#### Set Up

Loading Packages and metadata

```{r setup}
#load packages and set up colors for plotting
source("scripts/00_setup.R")

```

#### This script will:

-   Intake water chemistry data from RMRS partners
-   Remove duplicates and blanks used by RMRS for QAQC
-   Add time data missing from RMRS dataset
-   Add location metadata

```{r raw_data}
#Change to name of newest file
newest_raw_file <- "ROSS_20251114.csv"

newest_meas_file <- 'ROSS_field_meas_upto_2025-11-14.csv'

```

```{r cleaning}
#read most recent file
water_chemistry_import <- read_csv(here("data", "raw", newest_raw_file), show_col_types = FALSE)%>%

  #add FCW number to FCW samples (for IDs in reservoir study)
  mutate(FCW_number = case_when(str_detect(LabID, "FCW") ~  as.numeric(gsub('FCW ', '', LabID)),
                              TRUE ~ NA),
         ROSS_number = case_when(str_detect(LabID, "ROSS") ~  as.numeric(gsub('ROSS ', '', LabID)),
                              TRUE ~ NA),
         #parse date
         Date = as.Date(Date, format = "%d-%b-%y"),
         #get year
          Year =  year(Date))%>%
  rename(site_code = SiteLabel)%>%
  #omitting prospect samples with likely bottle contam (High values which are outside of typical ranges for site)
  filter(!(site_code == "prospect" & (SC > 3000)))
  
```

# Adding data not given by RMRS

## Read in metadata

```{r meta}

# site_meta: Locations of each site, common name, location characteristics, WQ sensor owner, flow gauge information. Created by Sam Struthers
site_meta <- read_csv(here("data","metadata","location_metadata.csv"), show_col_types = FALSE) 

# ROSS field measurements archive: FCW number, Time (mst) and handheld probe values for all FCW samples (2023 prior)
old_field_meas <- read_csv(here("data","raw","ROSS_field_meas_archive.csv"), show_col_types = FALSE)

# more recent ROSS field measurements
#created from mWater field notes and imported by the script `scripts/generate_ross_field_meas.R`
new_field_meas <- read_csv(here("data", "raw", newest_meas_file), show_col_types = FALSE)%>%
  mutate(Date = as.Date(Date, format = "%m/%d/%y"), 
         DT_round = with_tz(DT_round, tzone = "MST"))

#buffer_sbs:Indices based on burn severity directly around the reservoir. Created by Katie Willi
buffer_sbs <- read_csv(here("data","metadata","sbs_buffer.csv"), show_col_types = FALSE) %>%
  mutate(Buffer_Level=((Unburned*0)+(V_Low*0.1)+(Low*0.4)+(Moderate*0.7)+(High*1))/(Unburned+V_Low+Low+Moderate+High))
# watershed_sbs: within each reservoirs watershed.Created by Katie Willi
watershed_sbs <- read_csv(here("data","metadata","sbs_watershed.csv"), show_col_types = FALSE) %>%
  mutate(Watershed_Level=((Unburned*0)+(V_Low*0.1)+(Low*0.4)+(Moderate*0.7)+(High*1))/(Unburned+V_Low+Low+Moderate+High))
#dist_from_pbd: distance (for upper sites) from PBD (mouth of canyon) using NHD flowline. 
#Created by Katie Willi using `scripts/distance_finder.R`
dist_from_pbd <- read_csv(here("data","metadata","distance_from_pbd.csv"), show_col_types = FALSE)

```

## Joining metadata to dataset

```{r join}
# add metadata and field measurements to water chemistry data
water_chemistry_joined <- water_chemistry_import%>%
  #Remove duplicate and blank samples that were used by RMRS for QA
  filter(SampleType %nin% c("BLANK", "DUPLICATE"))%>%
  #add location metadata created by Sam Struthers
  left_join(site_meta,by="site_code")%>%
  #add field measurements archive for all FCW samples (fewer notes were taken at this time). Compiled by Sam Struthers
  left_join(old_field_meas, by = "FCW_number")%>%
  #add more recent field measurements
      left_join(new_field_meas, by = c("Date", "site_code", "SampleType"))%>%
  #add buffer burn severity data created by Katie Willi
   left_join(select(buffer_sbs,c(watershed,Buffer_Level)),by="watershed") %>%
  #add watershed burn severity data created by Katie Willi
  left_join(select(watershed_sbs,c(site_code,Watershed_Level)),by="site_code")%>%
  # Add distance upstream of pbd site pulled from NHDplus by Katie Willi
  left_join(select(dist_from_pbd,c(site_code,distance_upstream_km)),by="site_code")%>%
  # Fill in missing field data with archived data or new measurements, and join into a single column
mutate(final_Field_DO_mgL = case_when( 
                                  !is.na(do_mgl) ~ do_mgl,
                                  !is.na(Field_DO_mgL) ~ Field_DO_mgL,
                                  !is.na(`ROSS Field_DO_mgL`) ~ `ROSS Field_DO_mgL`,
                                  TRUE ~ NA), 
      final_Field_Cond_µS_cm = case_when( 
                                  !is.na(cond_ms_cm) ~ cond_ms_cm,
                                  !is.na(`ROSS Field_Cond_µS/cm`) ~ `ROSS Field_Cond_µS/cm`,
                                  !is.na(`Field_Cond_µS/cm`) ~ `Field_Cond_µS/cm`,
                                  TRUE ~ NA),
          final_Field_Temp_C = case_when( !is.na(temp_c) ~ temp_c, 
                                          !is.na(`ROSS Field_Temp_C`) ~ `ROSS Field_Temp_C`,
                                          !is.na(Field_Temp_C) ~ Field_Temp_C,
                                  TRUE ~ NA), 
      final_time_mst = case_when( !is.na(`ROSS Time (MST)`) ~`ROSS Time (MST)`,
                                  !is.na(time_mst) ~ time_mst,
                                  !is.na(Time_mst) ~ Time_mst, 
                                  # `ROSS Time (MST)` is not accurate and should not be used
                                  TRUE ~ NA), 
      #Create a character version of date time in Mst
      DT_mst_char = case_when( !is.na(final_time_mst) ~ paste0(Date, " ", final_time_mst),
                                  TRUE ~ NA), 
      #convert to POSIXct object
      DT_mst = as.POSIXct(DT_mst_char, format = "%Y-%m-%d %H:%M", tz = "MST")
)
   

```

Do internal checks on data after this chunk.

- Check for duplicate names or samples
- Check for missing values from field measurements sheets
- Check for date or time errors

# Cleaning up the data

```{r cleaning}
# remove extra objects now that the data has been joined and tested
rm(new_field_meas, old_field_meas, site_meta, water_chemistry_import, newest_raw_file, buffer_sbs, dist_from_pbd, watershed_sbs, newest_meas_file)

water_chemistry_tidy <-  water_chemistry_joined %>%
  #remove extra columns
  select(-c(do_mgl, `ROSS Field_DO_mgL`, Field_DO_mgL, cond_ms_cm, `ROSS Field_Cond_µS/cm`,`Field_Cond_µS/cm`, temp_c, `ROSS Field_Temp_C`, Field_Temp_C,DT_round,time_mst, Time_mst,  
            # These columns are primarily used by RMRS for analysis and are not necessary for downstream analysis
            LabID,`ROSS Time (MST)`, `ROSS Turbidity`,Site_Name,  SampleType, Q_cfs, SiteDescr))%>%
  rename(time_mst = final_time_mst, 
         Field_DO_mgL = final_Field_DO_mgL, 
         Field_Cond_µS_cm = final_Field_Cond_µS_cm, 
         Field_Temp_C = final_Field_Temp_C)%>%
#arrange by date
  arrange(Date, DT_mst)%>%
  #clean up the order of columns
  select(site_code,Site, Date,DT_mst, Turbidity:SO4, Field_DO_mgL, Field_Cond_µS_cm, Field_Temp_C, Lat:flow_gauge_id, distance_upstream_km, Watershed_Level, Buffer_Level, FCW_number, ROSS_number, chla_volume_ml,Year, time_mst, DT_mst_char )%>%
  mutate( 
          #add data source column for data collected by ROSS (2021-2023) and Rhoades (2014-2023)
          data_source = case_when(Year >= 2021 ~ "ROSS",TRUE ~ "RHOADES"))

rm(water_chemistry_joined)
```

# Saving to CSV and RDS files

```{r saving}

# grab most recent date from chemistry dataset
up_to_date <- format(max(water_chemistry_tidy$Date), "%Y%m%d")
#write dataset to csv

file_name <- here("data","cleaned", paste0("CLP_chemistry_up_to_", up_to_date))

write_csv_arrow(select(water_chemistry_tidy, -DT_mst),paste0(file_name, ".csv") )

# Write to RDS file as well
write_rds(x = water_chemistry_tidy, file = paste0(file_name, ".rds"))

```

# Clean up 

```{r}
#Move old file to clean_archive

old_files <- grep(list.files(path=here("data","cleaned"), full.names = T), pattern=file_name, value = TRUE, invert = TRUE)
# copy to clean_archive
file.copy(from = old_files, to = here("data","clean_archive"), overwrite = TRUE)
# remove from cleaned
file.remove(old_files)

#clean up workspace
rm(water_chemistry_tidy, up_to_date, file_name, old_files)

```

## After running:

- Ensure new cleaned file is added to `data/cleaned` and that previous file is moved to `data/clean_archive`
- If new site is sampled, update maps in `scripts/sites_map.R` and ensure all locations are accurate before publication
